---
title: "1. Installation"
description: >
  Alternatives to install the orange R package. 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


### Development version using `devtools`

The simplest way to install the in-development version is via the `devtools` package. You just have to indicate that you want to install from the in-development `main` branch.

```{r, eval=FALSE}
devtools::install_github("adamkocsis/orange@main")
```

### Manual install

You can also the install the package manually with the following steps:

1. Clone the repository to your local hard drive.  
2. Make sure that you have the package dependencies.   
3. Open a terminal and navigate to the directory where you cloned. The `orange` directory should be visible from there.  
4. Run this line in the terminal  

```
R CMD INSTALL orange
```

- *If you see an error suggesting that `R` is not found, you have to add it to your `PATH` environmental variable.*  
- *If the R packages that `orange` depend on are not installed, you have to install them manually, or you will get an error.*

